A* with Perfect Potentials

### 2020-05-06 paper/ch_potentials.tex
add preliminary applications table, fix a lot of cites

Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
While flexible, A*'s performance crucially depends on the quality of the potential, also called heuristic, function.
Finding such a function that can be computed efficiently is difficult.
A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{CH,CCH,CRP,HIGHWAYHIERACHIES,MLD} which do not require such a function.
A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{geisberger}.
There is a lot of work that extends CH to handle more complex settings.
For example, in \cite{sabine, kobitzsch} multi-criteria optimization is studied.
In \cite{CCH}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in \cite{TD-CCH-preprint, TDCH, TDCRP}.
Specialized variants for electric vehicle routing are developped in \cite{moritz baum, charge}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.

While undoubtedly fast, hierarchical approaches tend to sacrifice flexibility to obtain speed.
In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
We are not the first to combine hierarchical approaches and A* \cite{core-alt, real, charge, bdsssw-chgds-08}.
However, past work mostly uses A* to accelerate hierarchical approaches and sacrifzies A*'s felxibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
The novel idea is to compute optimal A* potentials lazily.
The hierarchical approach is applied to the lower bound graph with respect to the edge costs.
As the hierarchical approaches are optimal, the resulting A* potential is the tightest possible.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
Figure~\ref{img:search-space} depicts an example  A* search with CH-Potentials.

@@

As detailed in the previous section, the reasons why flexibility is a necessary property of a routing engine are sheer endless.
However, a formal problem specification that encompasses all these options is complex and well beyond the scope of this paper.
For an example of such a formulation, we refer to \cite{sea2020paper} which covers a truck-specific application.
The algorithm presented in \cite{sea2020paper} uses a preliminary version of CH-Potentials.
For simplicity, we limit our exposition to the classical shortest path formulation.
To support algorithm extensions, we avoid two operations that are adverse to generalization.
These are:
(1) traversing an input edge in backward direction, and
(2) combining input weights.
Backward traversal is problematic as a weight can depend on the moment that an edge is entered, which is unknown in a backward search.
Combining input weights can be difficult, if the input weights are more complex objects than scalar values.
For example, when considering historic traffic patterns, the input weights are usually functions \cite{citation needed}.

@@

TODO: Dijkstra's algorithm \cite{Dijkstra's} can be used to solve the stated problem formalization \cite{Veit?}.

@@

Variations of this setup exist.
A very common assumption is that the query edge weights $w_q$ are known during the preprocessing phase.
In this setup, $w_\ell$ does not exist.
This allows for very fast query running times, however, changing $w$ requires rerunning the slow preprocessing phase.
This setup is therefore very unflexible.
Many hierarchical speedup techniques such as the original CH \cite{CH} use this setup.
Notable exceptions are \cite{MLD,CRP} and \cite{CCH} which use a third phase to change the query edge weights.

@@

A* star is a well-known extension to Dijkstra's algorithm.
It uses a potential function $\pi:V\rightarrow \mathbb{R}^+$.
The potential is often called heuristic.
A* is optimal, if the potential $\pi(v)$ is consistent \cite{Pearl, Judea (1984). Heuristics: Intelligent Search Strategies for Computer Problem Solving. Addison-Wesley. ISBN 0-201-05594-5.}.
In this paper, all considered potentials are derived from exact shortest path distances and are thus consistent.

@@

A CH is a two phase technique to efficiently compute optimal, shortest paths.
For a detailed exposition, we refer the interested reader to \cite{CCH, CH}.
In this section, we give a high-level overview of the aspects relevant to this paper.

@@

A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case as demonstrated in \cite{ch,punch,crp,cch,flowcutter}.

@@

This works, however, while one CH query is fast, answering one for every node in the $A^*$ search is slow.
Fortunatly, we can do better.
However, for this we need another component called Phast \cite{Phast}.

@@

Phast \cite{Phast} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogously to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH level from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all level are processed, the distances from all nodes towards $t$ is computed.
Phast is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
Phast's main advantage is that is easily be parallelized.
However, we will not consider parallelization in this paper.
We refer to \cite{Phast} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.

@@

In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
We do this using Tarjan's algorithm \cite{Tarjan's algorithm}.
For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
For nodes in the core, $a_v=v$.
We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
The basic CH-potential preprocessing step is only executed for $G_C$.

@@

Tims plan:

\begin{itemize}
\item Applications
	\begin{itemize}
  \item Table
  	\begin{itemize}
    \item columns: preprocessing time, running time, potential evals, relaxed arcs, shortest path distance / lower bound, dijkstra running time?
    \item rows: OSM Ger Mapbox Live, forbidden highways, forbidden tunnels, TD Ger06, TD Eur17, DIMACs Eur 1000 {Random|Fastest} Arcs {x2, x10}
		\end{itemize}
	\end{itemize}
\end{itemize}

>>>>

Dijkstra's algorithm \cite{d-ntpcg-59} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
While flexible, A*'s performance crucially depends on the quality of the potential, also called heuristic, function.
Finding such a function that can be computed efficiently is difficult.
A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{gssv-erlrn-12,dsw-cch-15,dgpw-crprn-13,ss-ehh-12} which do not require such a function.
A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{gssv-erlrn-12}.
There is a lot of work that extends CH to handle more complex settings.
For example, in \cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
In \cite{dsw-cch-15}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in \cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
Specialized variants for electric vehicle routing are developped in \cite{bdgwz-sfpcs-19}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.

While undoubtedly fast, hierarchical approaches tend to sacrifice flexibility to obtain speed.
In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
We are not the first to combine hierarchical approaches and A* \cite{bdsssw-chgds-10, gkw-blwr-07, bdgwz-sfpcs-19}.
However, past work mostly uses A* to accelerate hierarchical approaches and sacrifzies A*'s felxibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

Our algorithm is built upon ideas from Phast \cite{dgnw-phast-13}, a CH-extension.
The novel idea is to compute optimal A* potentials lazily.
The hierarchical approach is applied to the lower bound graph with respect to the edge costs.
As the hierarchical approaches are optimal, the resulting A* potential is the tightest possible.
With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.

@@

As detailed in the previous section, the reasons why flexibility is a necessary property of a routing engine are sheer endless.
However, a formal problem specification that encompasses all these options is complex and well beyond the scope of this paper.
For an example of such a formulation, we refer to \cite{kswz-erptd-20} which covers a truck-specific application.
The algorithm presented in \cite{kswz-erptd-20} uses a preliminary version of CH-Potentials.
For simplicity, we limit our exposition to the classical shortest path formulation.
To support algorithm extensions, we avoid two operations that are adverse to generalization.
These are:
(1) traversing an input edge in backward direction, and
(2) combining input weights.
Backward traversal is problematic as a weight can depend on the moment that an edge is entered, which is unknown in a backward search.
Combining input weights can be difficult, if the input weights are more complex objects than scalar values.
For example, when considering historic traffic patterns, the input weights are usually functions \cite{citation needed}.

@@

TODO: Dijkstra's algorithm \cite{d-ntpcg-59} can be used to solve the stated problem formalization \cite{Veit?}.

@@

Variations of this setup exist.
A very common assumption is that the query edge weights $w_q$ are known during the preprocessing phase.
In this setup, $w_\ell$ does not exist.
This allows for very fast query running times, however, changing $w$ requires rerunning the slow preprocessing phase.
This setup is therefore very unflexible.
Many hierarchical speedup techniques such as the original CH \cite{gssv-erlrn-12} use this setup.
Notable exceptions are \cite{dgpw-crprn-13} and \cite{dsw-cch-15} which use a third phase to change the query edge weights.

@@

A* star is a well-known extension to Dijkstra's algorithm.
It uses a potential function $\pi:V\rightarrow \mathbb{R}^+$.
The potential is often called heuristic.
A* is optimal, if the potential $\pi(v)$ is consistent \cite{p-hissc-84}.
In this paper, all considered potentials are derived from exact shortest path distances and are thus consistent.

@@

A CH is a two phase technique to efficiently compute optimal, shortest paths.
For a detailed exposition, we refer the interested reader to \cite{dsw-cch-15,gssv-erlrn-12}.
In this section, we give a high-level overview of the aspects relevant to this paper.

@@

A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case as demonstrated in \cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.

@@

This works, however, while one CH query is fast, answering one for every node in the $A^*$ search is slow.
Fortunatly, we can do better.
However, for this we need another component called Phast \cite{dgnw-phast-13}.

@@

Phast \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogously to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH level from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all level are processed, the distances from all nodes towards $t$ is computed.
Phast is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
Phast's main advantage is that is easily be parallelized.
However, we will not consider parallelization in this paper.
We refer to \cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.

@@

In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
For nodes in the core, $a_v=v$.
We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
The basic CH-potential preprocessing step is only executed for $G_C$.

@@

\begin{table}
\centering
% TODO dijkstra baseline, preprocessing
\caption{Performance for different applications.}\label{tab:applications}
\begin{tabular}{llrrrrr}
\toprule
        &                     &   Running &     Pot. Eval. &             Relaxed & Avg. length & Affected \\
        &                     & time [ms] & $[\cdot 10^3]$ & arcs $[\cdot 10^3]$ &    increase &     [\%] \\
\midrule
OSM Ger & No Tunnels          &      39.4 &         3874.1 &               290.6 &        1.74 &     83.5 \\
        & No Highways         &     338.5 &        30383.1 &              2555.9 &        6.24 &     95.0 \\
        & Mapbox Live         &      96.5 &         9310.9 &               733.7 &        1.12 &     95.0 \\
\addlinespace
DIMACs  & Fastest $\times 10$ &       1.0 &           69.2 &                 2.7 &        1.00 &     16.5 \\
Europe  & Fastest $\times 2$  &       1.0 &           68.9 &                 2.7 &        1.00 &     16.5 \\
        & Random $\times 10$  &       1.1 &           80.9 &                 3.0 &        1.00 &      1.0 \\
        & Random $\times 2$   &       1.0 &           76.7 &                 2.9 &        1.00 &      1.0 \\
\addlinespace
Ger06   & TD                  &       4.8 &          417.2 &                33.1 &        1.03 &     99.5 \\
Eur17   & TD                  &      98.5 &         5143.8 &               553.3 &        1.04 &     98.5 \\
\bottomrule
\end{tabular}
\end{table}
